### 2010-04-02 papers/icfp-2010-desc/paper_desc.tex
added split, tweaked the examples

To add inductive types to our type theory, we build a universe of
datatype \emph{descriptions} by implementing the signature below, with
codes mimicking the grammar of datatype declarations. We
can read a description $\V{D}:\Desc$ as an \emph{endofunctor}
on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
to be instantiated recursively.
%
\[\stk{
\begin{array}{l}
\Desc : \Set\\
\DUnit : \Desc\\
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\DIndx{(\Bhab{\V{D}}{\Desc})} : \Desc \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{X}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{X}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{X}}{\descop{\V{D}}{\V{X}}}
\end{array}
}\]
%
Descriptions are sequential structures, terminated by $\DUnit$, indicating
the empty tuple. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
where \(\V{D}\) descibes the rest of the node.

To give some intuition on this universe, we now turn to some
examples. For obvious pedagogical reasons, we will manually build the
descriptions. However, it should be clear that, in practice, these
definitions are automatically computed from an Agda-like $\data$
definition.

@@

Our first example is the natural numbers, or rather its pattern
functor. Our code is presented in the high-level expression language
of Section~\ref{sec:type-propagation}. The translation back to the raw
terms should not pose any difficulty. The code is the following:
%
\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ \DUnit \quad (\DIndx{\DUnit}) ]}
}\]
%
Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{}$ constructors. In
the $\NatZero$ case, we reach the end of the description. In the
$\NatSuc{\!}$ case, we attach a recursive argument and close the
description. In a more synthetic notation, we have described the
following functor:
%
\[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]

@@

With a small change to this definition, we obtain the pattern functor
for lists:
%
\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: \V{X} \mapsto \DSigma{(\EnumT{[ \ListNil, \ListCons{\!}{\!} ]})}
                           {[ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) ]}
}\]
%
The $\NatSuc{\!}$ constructor is turned into a proper
$\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
recursive argument. This code describes the following functor:
%
\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of binary trees:
%
\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: \V{X} \mapsto \DSigma{ & (\EnumT{[ \TreeLeaf, \TreeNode ]}) \\}
                           { & [ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{(\DIndx{\DUnit})}}) ]}
\end{array}
}\]
%
Again, we are one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we call it twice. The
interpretation of this code corresponds to the following functor:
%
\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \Prod \V{Z} \Prod \V{Z}     \]

@@

From the examples above, we observe that datatypes are defined by a
$\DSigma{\!}{\!}$ which first argument is a finite set of
constructors. The descriptions fitting into this pattern are called
\emph{tagged} description. Formally, we have:
%
\[
 \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
\]
%
This is a clear reminiscence of the sum-of-products style,
characterized by a finite sum of constructors. Every description can
be expressed in this style, using a singleton constructor if
necessary. Conversely, a tagged description can be turned into a
full-blown description. We denote $\toDesc{\V{TD}}$ the description
computed from the tagged description $\V{TD}$.

@@

For convenience, we have taken for granted the existence of $\Desc$,
as presented in Figure~\ref{fig:desc_universe}. Although it makes no
conceptual difference, this code should rather be read as a
specification. We promise the existence of such objects, satisfying
these typing rules. It will be the subject of
Section~\ref{sec:desc-levitate} to actually implement this
specification.

>>>>

To add inductive types to our type theory, we build a universe of
datatype \emph{descriptions} by implementing the signature below, with
codes mimicking the grammar of datatype declarations. We
can read a description $\V{D}:\Desc$ as a `pattern functor'
on $\Set$, with $\descop{\V{D}}{\!}$ its action on an object, \(\V{X}\), soon
to be instantiated recursively.
%
\[\stk{
\begin{array}{l}
\Desc : \Set\\
\DUnit : \Desc\\
\DSigma{(\Bhab{\V{S}}{\Set})}{(\Bhab{\V{D}}{\V{S} \To \Desc})} : \Desc\\
\DIndx{(\Bhab{\V{D}}{\Desc})} : \Desc \\
\end{array}
\smallskip\\
\begin{array}{l@{\V{Z}\:\mapsto\:}l}     
\multicolumn{2}{l}{\descop{\_\:}{} : \Desc \To \Set \To \Set} \\
 \descop{\DUnit}{} &  \Unit \\
 \descop{\DSigma{\V{S}}{\V{D}}}{} &
     \SIGMAS{\Bhab{\V{s}}{\V{S}}}{\descop{\V{D}\,\V{s}}{\!\V{Z}}}  \\
\descop{\DIndx{\V{D}}}{}  &  \TIMES{\V{Z}}{\descop{\V{D}}{\V{Z}}}
\end{array}
}\]
%
Descriptions are sequential structures, terminated by $\DUnit$, indicating
the empty tuple. To build
sigmas-of-sigmas, we define a $\DSigma{\!}{}$ code, interpreted as a
$\Sigma$-type. To request a recursive component, we have $\DIndx{\V{D}}$,
where \(\V{D}\) describes the rest of the node.

You may have noticed that we are a little coy about this presentation,
writing of `implementing a signature' without clarifying how. A viable
approach would simply be to extend the theory with constants for the
constructors and an operator for \(\descop{\V{D}}\). However, in
Section~\ref{sec:desc-levitate}, you will see what we actually do.  In
the meantime, let us first gain some intuition for its use by
developing some examples.

@@

We begin with the natural numbers, now working in the high-level
expression language of Section~\ref{sec:type-propagation}, exploiting
type propagation.
%
\[\stk{
\NatD : \Desc \\
\NatD \mapsto \DSigma{\EnumT{\sqr{\NatZero\: \NatSuc{\!} }}}
                     {\sqr{ \DUnit \quad (\DIndx{\DUnit}) }}
}\]
%
Let us explain its construction. First, we use $\DSigma{\!}{\!}$ to
give a choice between the $\NatZero$ and $\NatSuc{}$ constructors.
What follows depends on this choice, so we write the function
computing the rest of the description in tuple notation.  In the
$\NatZero$ case, we reach the end of the description. In the
$\NatSuc{\!}$ case, we attach one recursive argument and close the
description. Translating the \(\Sigma\) to a binary sum, we have
effectively described the functor:
%
\[    \NatD\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{Z}    \]
Correspondingly, we can see the injections to the sum:
\[
\sqr{\NatZero}:\descop{\NatD}Z \qquad
\sqr{\NatSuc\:z}:\descop{\NatD}Z \;\mbox{if}\;z:Z
\]

@@

With a small change to this definition, we obtain the pattern functor
for lists:
%
\[\stk{
\ListD : \Set \To \Desc \\
\ListD \: \V{X} \mapsto
 \DSigma{\EnumT{\sqr{ \ListNil\: \ListCons{\!}{\!} }}}
         {\sqr{ \DUnit \quad (\DSigma{\V{X}}{\LAM{\_} \DIndx{\DUnit}}) }}
}\]
%
The $\NatSuc{\!}$ constructor is turned into a proper
$\ListCons{\!}{\!}$, taking an argument in $\V{X}$ followed by a
recursive argument. This code describes the following functor:
%
\[    \ListD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} \V{X} \D{\ensuremath{\times}} \V{Z}     \]

@@

Finally, we are not limited to one recursive argument. This is
demonstrated by our description of node-labelled binary trees:
%
\[\stk{
\TreeD : \Set \To \Desc \\
\begin{array}{@{}ll}
\TreeD \: \V{X} \mapsto \DSigma{ & \EnumT{\sqr{ \TreeLeaf, \TreeNode }} \\}
  { & \sqr{ \DUnit \quad
       ( \DIndx{(\DSigma{\V{X}}{\LAM{\_}\DIndx{\DUnit}})})} }
\end{array}
}\]
%
Again, we are one evolutionary step away from $\ListD$. However,
instead of a single call to the induction code, we add another. The
interpretation of this code corresponds to the following functor:
%
\[    \TreeD\: \V{X}\: \V{Z} \mapsto \Unit \mathop{\D{+}} 
          \V{Z} \Prod \V{X}  \Prod \V{Z}     \]

@@

From the examples above, we observe that datatypes are defined by a
$\DSigma{\!}{\!}$ whose first argument enumerates the constructors. We
call codes fitting this pattern \emph{tagged} descriptions. Again,
this is a clear reminder of the sum-of-products style. Every
description can be forced into this style with a singleton constructor
if necessary. We characterise tagged descriptions thus:
\[\stk{
 \TagDesc : \Set \\
 \TagDesc \mapsto \SIGMA{\V{E}}{\EnumU} (\spi{\V{E}}{(\LAM{\_} \Desc)})
\smallskip\\
\toDesc{\!} : \TagDesc\To\Desc \\
\toDesc{\!} \mapsto
\spl{\LAM{\V{E}}\LAM{\V{D}}
\DSigma{\EnumT{\V{E}}}{(\F{switch}\:\V{E}\:(\LAM{\_}\Desc)\:\V{D})}}
}\]

@@
